---
title: "02_Data_manipulation"
author: "Elizabeth Bowman"
date: "2/16/2020"
output: html_document
---

# Common dplyr functions

## select(): subset columns
The first argument in dplyr function should be the data frame name.
```{r}
select(surveys, plot_id, species_id, weight)
```

'-' in front of a variable excludes that variable from the output data frame

```{r}
select(surveys, -record_id, -species_id)
```

## filter(): subset rows on conditions
```{r}
filter(surveys, year == 1995)
```

Now, I would like to put filter and select together,but first we need to learn about 'pipes'.

## What is a 'pipe'?
A 'pipe' moves an object foward from one function or command to the next. It is intended to a) increase the readability and maintainability of code and b) decrease code development time.

For example:
2*(3*(4 + 4))

4 + 4 %>% 
  *3 %>%
  *2

```{r}
car_data <- transform(aggregate(. ~ cyl,
                    data = subset(mtcars, hp > 100),
                    FUN = function(x) round(mean(x,digits = 2))),
          kpl = mpg * 0.4251)
```

```{r}
mtcars %>%
  subset(hp > 100) %>%
  aggregate(. ~ cyl, data = ., FUN = . %>% mean %>% round(2)) %>%
  transform(kpl = mpg %>% multiply_by(0.4251)) %>%
  print
```

You are using the 'pipes' to progressively build on previous functions. See here[linked phrase](http://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html) for more information on 'pipes'.

Now to put what we've learned together.
```{r}
surveys %>%
  filter(year == 1995) %>%
  select(species_id, sex, weight) 
```

You can check each step individually if you need by running each line one-by-one. This makes it easy to see where an error is occurring in your code.

# Activity
Using pipes, subset the surveys data to include animals that weigh less than 5 grams and only keep columns species_id, sex, and weight.
Hints: 
* for less than use the symbol <
* remember: the weight column is in grams, so you do not need to state that

Answer:
```{r}
surveys %>% 
  filter(weight < 5) %>%
  select(species_id, sex, weight) -> surveys.small
head(surveys.small)
```

## mutate(): create new columns by using information other than columns
This is a very useful function when you want to create a new column based on the values in an already existing column, i.e. unit conversions or a ratio value fom two other columns.
The weight in our surveys data frame is in grams. Let's convert it to kg.
```{r}
surveys %>% 
  mutate(weight_kg = weight/1000)
```
Let's add another column in pounds. 
```{r}
surveys %>% 
  mutate(weight_kg = weight/1000,
         weight_lbs = weight_kg *2.2) %>%
  tail()
```

# Activity
Create a new data frame which contains
* only the species_id column and hindfoot_half column,
* a new column called hindfoot_half which contains values that are half the hindfoot_length values, 
* and only hindfoot_half values less than 30.

Answer:
```{r}
surveys %>%
  mutate(hindfoot_half = hindfoot_length/2) %>%
  filter(hindfoot_half < 30) %>%
  select(species_id, hindfoot_half) -> surveys.hfh
```

## group_by() and summarize(): create summary statistics on group data
A common approach used in data analysis is splitting data and 



