---
title: "02_Data_manipulation"
author: "Elizabeth Bowman"
date: "2/16/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(magrittr)

surveys <- read.csv('~/Desktop/TidyverseWS/portal_data_joined.csv')
```
# Common dplyr functions

## select(): subset columns
The first argument in dplyr function should be the data frame name.
```{r select, eval = F}
select(surveys, plot_id, species_id, weight)
```

'-' in front of a variable excludes that variable from the output data frame

```{r neg select, eval = F}
select(surveys, -record_id, -species_id)
```

## filter(): subset rows on conditions
```{r filter, eval= F}
filter(surveys, year == 1995)
```

Now, I would like to put filter and select together,but first we need to learn about 'pipes'.

## What is a 'pipe'?
A 'pipe' moves an object foward from one function or command to the next. It is intended to a) increase the readability and maintainability of code and b) decrease code development time.

For example:
2*(3*(4 + 4))

4 + 4 %>% 
  *3 %>%
  *2

```{r example2 base}
car_data <- transform(aggregate(. ~ cyl,
                    data = subset(mtcars, hp > 100),
                    FUN = function(x) round(mean(x,digits = 2))),
          kpl = mpg * 0.4251)
head(car_data)
```

```{r example2 tidyr}
mtcars %>%
  subset(hp > 100) %>%
  aggregate(. ~ cyl, data = ., FUN = . %>% mean %>% round(2)) %>%
  transform(kpl = mpg %>% multiply_by(0.4251)) %>%
  head()
```

You are using the 'pipes' to progressively build on previous functions. See here[linked phrase](http://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html) for more information on 'pipes'.

Now to put what we've learned together.
```{r year95}
surveys %>%
  filter(year == 1995) %>%
  select(species_id, sex, weight)  %>%
  head()
```

You can check each step individually if you need by running each line one-by-one. This makes it easy to see where an error is occurring in your code.

# Activity 1
Using pipes, 
1. subset the surveys data to include animals that weigh less than 5 grams and
2. only keep columns species_id, sex, and weight.

Hints: 

* For 'less than', use the symbol <
* Remember: the weight column is in grams, so you do not need to state that

Answer:
```{r Activity1}
surveys %>% 
  filter(weight < 5) %>%
  select(species_id, sex, weight) -> surveys.small
head(surveys.small)
```

## mutate(): create new columns by using information other than columns
This is a very useful function when you want to create a new column based on the values in an already existing column, i.e. unit conversions or a ratio value fom two other columns.
The weight in our surveys data frame is in grams. Let's convert it to kg.
```{r mutate kg}
surveys %>% 
  mutate(weight_kg = weight/1000) %>%
  tail()
```
Let's add another column in pounds. 
```{r mutate kg lb}
surveys %>% 
  mutate(weight_kg = weight/1000,
         weight_lbs = weight_kg *2.2) %>%
  tail()
```

# Activity 2
Create a new data frame which contains
1. only the species_id column and hindfoot_half column,
2. a new column called hindfoot_half which contains values that are half the hindfoot_length values, and
3. only hindfoot_half values less than 30.

Answer:
```{r Activity2}
surveys %>%
  mutate(hindfoot_half = hindfoot_length/2) %>%
  filter(hindfoot_half < 30) %>%
  select(species_id, hindfoot_half) -> surveys.hfh
```

## group_by() and summarize(): create summary statistics on group data
A common approach used in data analysis is splitting data into groups based on a categorical variable, applying an analysis to each group, and then combining the results (split-apply-combine paradigm). dplyr uses the group_by() function to achiee this.

group_by() is used often with summarize(). summerize() collapses each group into a single row summary statistic for each group (e.g. average, minimum).

Let's calculate the mean weight by sex.
```{r group summarize}
surveys %>%
  group_by(sex) %>%
  summarize(mean_weight = mean(weight,na.rm = T))
```

We can also group by multiple columns.
```{r group mult col}
surveys %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight,na.rm = T))
```

If you notice, the mean_weight column has NaN (Not a Number). Let's remove the missing values for weight before calculating the mean.
```{r remove NaN}
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight,na.rm = T))
```

## arrange(): sort results
Let's add another variable, minimum weight or min_weight, to the summary statistics table and arrange the groups by mean weight.
```{r arrange}
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight,na.rm = T),
            min_weight = min(weight)) %>%
  arrange(mean_weight)
```

Arranging in desciending order (highest to lowest)
```{r arrange desc}
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight,na.rm = T),
            min_weight = min(weight)) %>%
  arrange(desc(mean_weight))
```

## count(): count discrete values
When working with ecological data, we often want to know the number of observations found for each factor or group of factors. Think of a site x species matrix.
Let's count the number of observations for each sex.
```{r count}
surveys %>%
  count(sex, sort = T)
```
'sort = T' sorts the count number in descending order.

If you would like to combine values, you can list them such as here where I am counting the numbers of each sex from each species.
```{r count by species}
surveys %>%
  count(sex, species, sort = T)
```

# Activity 3
Create a table with a count of the the number of species (species_id) within each plot (plot_id).

Answer:
```{r activity3}
surveys %>%
  count(plot_id, species_id)
```

## Reshaping with spread and gather
Remember the rules of a tidy dataset?
1. Each variables has its own columns.
2. Each observation has its own rows.
3. Each value must has its own cell.
The fourth rule is 'Each type of observational unit forms a table'. If we are interested in comparing the mean weight of each species between plots, we would need to create a table where each row is is values of variables association with each plot.

For instance, we want to know the mean weight of each genus in each plot. Let's create table that shows that. 
```{r mean wt}
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(genus, plot_id) %>%
  summarize(mean_weight = mean(weight)) -> surveys_mean
head(surveys_mean)
```

This almost gets us there, but we need to reshape the data with genera as the columns and plot_ids as the rows. To do this, we need spread() to transform the data, i.e. take a column and spread it out across rows.
```{r spread}
surveys_mean %>%
  spread(key = genus, value = mean_weight, fill = 0) -> surveys_spread
head(surveys_spread)
```

# Activity 4
The output from **Activity 3** is almost a site x species matrix. The only difference is that in a site by species matrix you have sites as the rows and species as the columns. Use the code from **Activity 3** to create a site x species matrix with the spread() function.

Answer:
```{r activity4}
surveys %>%
  count(plot_id, species_id) %>%
  spread(key = species_id, value = n, fill = 0) -> surveys_specobs
```

The opposing function gather() takes data spread out across multiple rows and gathers it into a single column. Let's transform surveys_spread back using the gather() function.
```{r gather}
surveys_spread %>%
  gather(key = genus, value = mean_weight, -plot_id) -> surveys_gather
```

We created a key called genus and value called mean_weight, i.e. we are creating two columns. The '-plot_id' signifies that the function will use all columns except plot_id for the key variable.

## Exporting data
```{r write csv}
write_csv(surveys_spread, pat = 'sitexspecies.csv')
```
