---
title: "03_Data_visualization"
author: "Elizabeth Bowman"
date: "2/16/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(magrittr)
surveys <- read.csv('~/Desktop/TidyverseWS/portal_data_joined.csv')
```

# GGplot2
ggplot graphics are built step-by-step by adding elements or layers. The basic template for ggplot is
```{r ggplot basic, eval = F}
ggplot(data = <data>,
       mapping = aes(x = <x.variable>,
                     y = <y.variable>)) +
  <geom_function>
```

* data specifies the data frame from which variables will be taken
* mapping is used to specify the variables that will be plotted 
* geoms are the graphical representation of the data (e.g. points, box and whisker, lines)
  * geom_point()
  * geom_boxplot()
  * geom_line()
* Elements or layers are added successively to the plot by using the + operator. You can add any number of parameters to your plot to personalize it to your specifications. I'll show you some aspects of this later. Let's start with a basic plot

## Distribution plot
Let's try to plot a distribution of weight from our animal survey data to assess our data.
```{r barplot weight}
ggplot(data = surveys,
       mapping = aes(x = weight)) +
  geom_bar()
```
Right away, we can see that our data is right skewed.

## Box and whisker plot
If we want to look at weight (continuous variable) as a function of animal genus (categorical variable), we can use geom_boxplot().
```{r boxwhisker}
ggplot(data = surveys,
       mapping = aes(x = genus,
                     y = weight)) +
  geom_boxplot()
```

Let's add some layers by fixing the label titles and changing the direction of the x axis labels.
```{r boxwhisker axes}
ggplot(data = surveys,
       mapping = aes(x = genus,
                     y = weight)) +
  geom_boxplot() +
  xlab('Genus') +
  ylab('Weight (g)') +
  theme(axis.text.x=element_text(angle = -90, hjust = 0))
```

## Scatter plot
Another basic plot, is the scatter plot. To produce a scatter plot, your x and y variable both need to be continuous. To plot this, we will use the geom_point() function instead of geom_boxplot().

For an example, let's plot weight as a function of hindfoot_length.
```{r scatter}
ggplot(data = surveys,
       mapping = aes(x = hindfoot_length,
                     y = weight)) +
  geom_point()
```

# Activity 1

Let's create a new table in which we

* filter values of weight and hindfoot_length that are NA
* filter year values that are greater than 1999.
* group data by sex and year
*select variables sex, year, hindfoot_length, and weight

Using these parameters, create a new data frame called plot.df
```{r Activity1}
surveys %>%
  filter(!is.na(weight), !is.na(hindfoot_length), year > 1999) %>%
  group_by(sex, year) %>%
  select(sex, year, hindfoot_length, weight) -> plot.df
```

Let's build a scatter plot, in which we a) plot weight as a function of hindfoot_length, b) divide plots out by sex and by year, c) include linear regressions in each plot, and d) modify the x and y axis label.

### Base scatter plot
```{r Activity1 base}
ggplot(plot.df, aes(x = hindfoot_length,
                    y = weight)) +
  geom_point(size = 2)
```

### Divide plots by sex and year
```{r Activity1 division}
ggplot(plot.df, aes(x = hindfoot_length,
                    y = weight)) +
  geom_point(size = 2) +
  facet_grid(vars(sex), vars(year))
```

### Add linear regression lines
```{r Activity1 Plot}
ggplot(plot.df, aes(x = hindfoot_length,
                    y = weight)) +
  geom_point(size = 2) +
  facet_grid(vars(sex), vars(year)) +
  geom_smooth(method = 'lm', se = F, color = 'darkgrey')
```

### Modify labels for x and y axes
```{r Activity1 Plot}
ggplot(plot.df, aes(x = hindfoot_length,
                    y = weight)) +
  geom_point(size = 2) +
  facet_grid(vars(sex), vars(year)) +
  geom_smooth(method = 'lm', se = F, color = 'darkgrey') +
  ylab('Weight (g)') +
  xlab('Hindfood length (mm)')
```

